\documentclass{article}

\usepackage{listings}
\usepackage{graphicx}

\title{Using Maxima to write a \LaTeX~ article}
\author{Richel Bilderbeek}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
This article is created from within the CAS program Maxima
and converted to PDF by using FOSS tools only, to assess the viability of this pipeline.As a testcase, an exercise of \cite{case2000} (page 69) is used.
\end{abstract}

\section{Introduction}

\LaTeX~is commonly used for writing publishable scientific articles\cite{gaudeul2006}.
Algebraic manipulations can be done by a CAS, for example Maxima, Maple or Mathematica.
Maxima is the only free and open-source program, and it is the oldest free and open-source computer algebra system, with development started in 1967 (as Macsyma) or 1982 (as MAXIMA).
This article is an example of writing a \LaTeX~ article within Maxima

Writing \LaTeX~ is slower and introduces more errors 
then using Microsoft Word as a text editor \cite{knauff&nejasmic2014}.
In this article, however, the text of a document is generated. \LaTeX~ can
parse that text to create a document, like a Word .docx could do as well,
with less markup (XML) added. Additionally, all tools in the current pipeline
are FOSS, so anyone with an internet connection can access these without
paying licensing costs.

Maxima its capabilities is tested by doing an exercise of \cite{case2000} at page 69.

\section{Exercise}

\begin{table}[here]
  \centering
  \begin{tabular}{ | r | l | }
    \hline
    symbol & description \\
    \hline
    $I$ & Indentity matrix \\
    $L$ & Leslie matrix \\
    $M$ & Leslie matrix with $\lambda$ subtracted at diagonal \\
    $x$ & population density vector \\
    $Z$ & Vector filled with zeroes \\
    $\lambda$ & population growth rate \\
    \hline
  \end{tabular}
  \caption{Definitions}
  \label{table:table_definition}
\end{table}

The equations to solve are:
\begin{equation}
x\,L=x\,\lambda\label{eq:eq3_21}
\end{equation}
\begin{equation}
x\,M=Z\label{eq:eq3_23}
\end{equation}

Equation \ref{eq:eq3_21} equals equation 3.21 of \cite{case2000} (or P2.4a, page 237, in \cite{otto&day2007})
Equation \ref{eq:eq3_23} equals equation 3.23 of \cite{case2000} 
(for definitions see table \ref{table:table_definition} on page \pageref{table:table_definition}).
The Leslie matrix, L,  given in this exercise is:

\begin{equation}
L=\pmatrix{1.0&1.0\cr 0.8&0.8\cr }\end{equation}

The Leslie matrix, L, used:

\begin{equation}
L=\pmatrix{1.0&1.0\cr 0.8&0.8\cr }\end{equation}


Note that this matrix has 2 age classes.
The simplifies equation \ref{eq:eq3_21} to:

\begin{equation}
\pmatrix{1.0\,x&1.0\,x\cr 0.8\,x&0.8\,x\cr }=x\,\lambda\end{equation}

Solving equation \ref{eq:eq3_21} can be done with equation \ref{eq:eq3_24}:

\begin{equation}
det(L - \lambda*I) = Z
\label{eq:eq3_24}
\end{equation}

Equation \ref{eq:eq3_24} equals equation 3.24 of \cite{case2000} (also: P2.4b at page 237, or 10.3.2 at page 406, in \cite{otto&day2007}) ,
where I is the identity matrix:
\begin{equation}
I=\pmatrix{1&0\cr 0&1\cr }\end{equation}
And Z is the vector of zeroes:
\begin{equation}
Z=\pmatrix{0\cr 0\cr }\end{equation}

This simplifies equation \ref{eq:eq3_24} to:

\begin{equation}
{\it det}\left(\pmatrix{1.0-\lambda&1.0\cr 0.8&0.8-\lambda\cr }\right)=\pmatrix{0\cr 0\cr }\end{equation}


The determinant of that matrix ($M$), is:
\begin{equation}
det(\pmatrix{1.0-\lambda&1.0\cr 0.8&0.8-\lambda\cr })=\left(0.8-\lambda\right)\,\left(1.0-\lambda\right)-0.8
\end{equation}

Solving $M=0$, the $\lambda$s found are:

\begin{equation}
\left[ \lambda={{9}\over{5}} , \lambda=0 \right] \end{equation}

There is one stable population structure, $\lambda=0$, which is
denotes an extinct population.
Here I focus on the more interesting value,
where $\lambda=9/5$.
This lambda is called the dominant eigenvalue, which equals the ultimate population growth.

Put $\lambda$ in M, this results in:
\begin{equation}
M=\pmatrix{-0.8&1.0\cr 0.8&-1.0\cr }\end{equation}

Now we can solve the $x$ of equation \ref{eq:eq3_23} (equals 3.23 from \cite{case2000}),
which was this:

\begin{equation}
x\,M=Z\end{equation}

This equation is unsolvable, unless we assign a value to an element of $x$.
Here, I put $1.0$ as the initial value of $x$ its first element.
(it will be rescaled later):

\begin{equation}
x = \pmatrix{1.0\cr {\it x_2}\cr }\end{equation}
Putting this $x$ in equation \ref{eq:eq3_23}:
\begin{equation}
M * x = \pmatrix{1.0\,{\it x_2}-0.8\cr 0.8-1.0\,{\it x_2}\cr }= \pmatrix{0\cr 0\cr }\end{equation}
As our matrix has two rows, there are two equations that can be solved:
Solving the upper, results in $x2=4/5$.
This results in an $x$ of:
\begin{equation}
x = \pmatrix{1.0\cr {{4}\over{5}}\cr }\end{equation}
$x$ must be rescaled so that its sum equals $1.0$.
$x$ its current sum is $1.8$, so dividing all elements by it, results in an $x$ of:
\begin{equation}
x = \pmatrix{0.55555555555555\cr 0.44444444444444\cr }\end{equation}
\section{Conclusion}

For this Leslie matrix:
\begin{equation}
\pmatrix{1.0&1.0\cr 0.8&0.8\cr }\end{equation}

The dominant eigenvalue, $\lambda$, is:
\\\\
\begin{equation}
\lambda = 9/5\end{equation}
\\

The stable population size distribution is:

\begin{equation}
x = \pmatrix{0.55555555555555\cr 0.44444444444444\cr }\end{equation}

\section{Discussion}

Writing \LaTeX~within Maxima can be done, but it is a bit cumbersome:
Maxima does not know \LaTeX~syntax and just creates contextless strings,
which might not be compilable by \LaTeX.
However, because the script does create a .tex file,
this file can be inspected easily with a \LaTeX~tool like texmaker.

\begin{thebibliography}{9}

\bibitem{case2000}
  Case, Ted J.
  2000
  An illustrated guide to theoretical ecology.

\bibitem{gaudeul2006}
  Gaudeul, A.
  2006
  Do Open Source Developers Respond to Competition?: The (La)TeX Case Study.
  Available at SSRN: http://ssrn.com/abstract=908946 or http://dx.doi.org/10.2139/ssrn.908946

\bibitem{knauff&nejasmic2014}
  Knauff, M. and Nejasmic, J.
  December 19, 2014
  An Efficiency Comparison of Document Preparation Systems Used in Academic Research and Development.
  PLoS ONE 9(12): e115069. doi: 10.1371/journal.pone.0115069

\bibitem{otto&day2007}
  Otto, Sarah P. and Day, T.
  2007
  A biologist's guide to mathematical modeling in ecology and evolution.
  ISBN-13: 978-0-691-12344-8

\end{thebibliography}

\appendix

\section{Script file}

\lstinputlisting[language=C++,showstringspaces=false,breaklines=true,frame=single]{case_2000_69.sh}

\section{Maxima file}

\lstinputlisting[language=C++,showstringspaces=false,breaklines=true,frame=single]{case_2000_69.txt}

\section{\LaTeX~file}

\lstinputlisting[language=tex,showstringspaces=false,breaklines=true,frame=single]{case_2000_69_output.tex}

\end{document}
